Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-29 manuscript/manuscript.Rmd
added equations for V(D) and V(C)

We estimate these using a form of variance decomposition. First, to measure stable interpersonal differences, we calculate the sum of the squared residuals derived from assuming that each person spends the entire survey at the estimated midpoint of their LCAM trajectory.[^2] We then divide this value by the total sum of squares and subtract this number from from 1, producing a value we denote as $V(D)$. This is the proportion of total variance attributable to interpersonal differences. (This number is equivalent to the conditional $R^2$ value generated by a model with only random intercepts for all respondents.)

Second, we calculate another sum of squared residuals generated by allowing the predictions to vary by estimated LCAM trajectory. We divide this number by the total sum of squares and subtract it from 1. We then subtract $V(D)$ from this value. This gives us a value we denote as $V(C)$, the additional proportion of variance accounted for by intrapersonal change.

[^2]: We use the midpoint estimate, rather than the value at the first wave, because this is the best measure of "baseline" available under the assumption that the person does not change. The first wave measure alone contains an unknown amount of measurement error and transient fluctuations.

These two numbers together reflect the total systematic variance of the LCAM that can be attributed to either differences between people or linear intrapersonal change over the course of the panel.[^3] Our principal outcome of interest is the proportion of systematic variance accounted for by intrapersonal change, which we calculate as

[^3]: This does not mean that other factors might not explain additional variance if they were included in the model. For example, peoples' responses about religious beliefs and participation might systematically vary by the day of the week or month in which they are surveyed. This variation is moved into the "random" or non-systematic component of the model for us.

@@

We can think of $\omega$ as a type of Intraclass Correlation Coefficient focused only on the systematic components of the model while ignoring residual variance that might be attributable to measurement error or non-persisting change [see also the variance decomposition of period and cohort by @vaisey2016]. The higher this proportion is, the more systematic variance in responses is attributable to intrapersonal change during the panel. From now on, we use $\omega$ as a placeholder for *systematic variance attributable to intrapersonal change*.

Because this quotient summarizes variance and therefore has no natural referent, it is hard to say at which values we observe "a lot" of intrapersonal change. But summarizing questions in this way allows us to compare the relative importance of our two broad processes across a range of questions that might have different levels of non-systematic variance. This gives us a sense of the relative prevalence of these two processes for each question, which is not achievable under the tournament of models approach. This also allows us to compare which kinds of questions show more or less intrapersonal change relative to interpersonal differences. While the tournament of models approach allowed researchers to say whether questions in general showed *any* evidence of updating, this approach allows us to make more specific claims about the prevalence of these theoretical processes for particular questions relative to each other.

>>>>

We estimate these using a form of variance decomposition. First, to measure stable interpersonal differences, we calculate $V(D)$ according to the equation below.[^2] 
$$
V(D) = 1 - \frac{\sum_{i=1}^N \sum_{t=1}^T [y_{it} - (\hat{\beta}_0 + \hat{\alpha}_{0i} + (\hat{\beta}_1 + \hat{\alpha}_{1i}) \overline{\text{year}}_{it})]^2}{\sum_{i=1}^N \sum_{t=1}^T [y_{it} - \bar{y}_{it}]^2}
$$

[^2]: We use the model estimate from the midpoint year, rather than the value at the first wave, because this is the best measure of "baseline" available under the assumption that the person does not change. The first wave measure alone contains an unknown amount of measurement error and transient fluctuations.

This gives us the proportion of total variance in an item attributable to stable interpersonal differences.

To calculate the second component, which is the variance attributable to systematic intrapersonal change, we calculate $V(C)$ as follows:

$$
V(C) = 1 - \frac{\sum_{i=1}^N \sum_{t=1}^T [y_{it} - (\hat{\beta}_0 + \hat{\alpha}_{0i} + (\hat{\beta}_1 + \hat{\alpha}_{1i}) \text{year}_{it})]^2}{\sum_{i=1}^N \sum_{t=1}^T [y_{it} - \bar{y}_{it}]^2} - V(D)
$$

This is the incremental proportion of variance accounted for when we allow the model predictions to change over time for each person.

The third component, which is the proportion of variance attributable to measurement error or transient fluctuations, is what's left, such that these three numbers sum to 1. As we discussed above, this component is not of core interest here.

Together, $V(D)$ and $V(C)$ reflect the total systematic variance of the outcome that the LCAM can attribute to either differences between people or linear intrapersonal change over the course of the panel. Our principal measure of interest is the proportion of systematic variance accounted for by intrapersonal change, which we calculate as

@@

We can think of $\omega$ as a type of Intraclass Correlation Coefficient focused only on the systematic components of the model while ignoring residual variance that might be attributable to measurement error or non-persisting change [see also the variance decomposition of period and cohort by @vaisey2016]. The higher this proportion is, the more systematic variance in responses is attributable to intrapersonal change during the panel. From now on, we use $\omega$ to stand for *the proportion of systematic variance attributable to intrapersonal change*.

Because this quotient summarizes variance proportions and therefore has no natural referent, it is hard to say at which values we observe "a lot" of intrapersonal change. But summarizing questions in this way allows us to compare the relative importance of our two broad processes across a range of questions that might have different levels of non-systematic variance. This gives us a sense of the relative prevalence of these two processes for each question, which is not achievable under the tournament of models approach. This also allows us to compare which kinds of questions show more or less intrapersonal change relative to interpersonal differences. While the tournament of models approach allowed researchers to say whether questions in general showed *any* evidence of updating, this approach allows us to make more specific claims about the prevalence of these theoretical processes for particular questions relative to each other.

### 2023-09-29 manuscript/manuscript.tex
added equations for V(D) and V(C)

We estimate these using a form of variance decomposition. First, to
measure stable interpersonal differences, we calculate the sum of the
squared residuals derived from assuming that each person spends the
entire survey at the estimated midpoint of their LCAM
trajectory.\footnote{We use the midpoint estimate, rather than the value
  at the first wave, because this is the best measure of ``baseline''
  available under the assumption that the person does not change. The
  first wave measure alone contains an unknown amount of measurement
  error and transient fluctuations.} We then divide this value by the
total sum of squares and subtract this number from from 1, producing a
value we denote as \(V(D)\). This is the proportion of total variance
attributable to interpersonal differences. (This number is equivalent to
the conditional \(R^2\) value generated by a model with only random
intercepts for all respondents.)

Second, we calculate another sum of squared residuals generated by
allowing the predictions to vary by estimated LCAM trajectory. We divide
this number by the total sum of squares and subtract it from 1. We then
subtract \(V(D)\) from this value. This gives us a value we denote as
\(V(C)\), the additional proportion of variance accounted for by
intrapersonal change.

These two numbers together reflect the total systematic variance of the
LCAM that can be attributed to either differences between people or
linear intrapersonal change over the course of the panel.\footnote{This
  does not mean that other factors might not explain additional variance
  if they were included in the model. For example, peoples' responses
  about religious beliefs and participation might systematically vary by
  the day of the week or month in which they are surveyed. This
  variation is moved into the ``random'' or non-systematic component of
  the model for us.} Our principal outcome of interest is the proportion
of systematic variance accounted for by intrapersonal change, which we
calculate as

@@

We can think of \(\omega\) as a type of Intraclass Correlation
Coefficient focused only on the systematic components of the model while
ignoring residual variance that might be attributable to measurement
error or non-persisting change (see also the variance decomposition of
period and cohort by Vaisey and Lizardo 2016). The higher this
proportion is, the more systematic variance in responses is attributable
to intrapersonal change during the panel. From now on, we use \(\omega\)
as a placeholder for \emph{systematic variance attributable to
intrapersonal change}.

Because this quotient summarizes variance and therefore has no natural
referent, it is hard to say at which values we observe ``a lot'' of
intrapersonal change. But summarizing questions in this way allows us to
compare the relative importance of our two broad processes across a
range of questions that might have different levels of non-systematic
variance. This gives us a sense of the relative prevalence of these two
processes for each question, which is not achievable under the
tournament of models approach. This also allows us to compare which
kinds of questions show more or less intrapersonal change relative to
interpersonal differences. While the tournament of models approach
allowed researchers to say whether questions in general showed
\emph{any} evidence of updating, this approach allows us to make more
specific claims about the prevalence of these theoretical processes for
particular questions relative to each other.

>>>>

We estimate these using a form of variance decomposition. First, to
measure stable interpersonal differences, we calculate \(V(D)\)
according to the equation below.\footnote{We use the model estimate from
  the midpoint year, rather than the value at the first wave, because
  this is the best measure of ``baseline'' available under the
  assumption that the person does not change. The first wave measure
  alone contains an unknown amount of measurement error and transient
  fluctuations.} \[
V(D) = 1 - \frac{\sum_{i=1}^N \sum_{t=1}^T [y_{it} - (\hat{\beta}_0 + \hat{\alpha}_{0i} + (\hat{\beta}_1 + \hat{\alpha}_{1i}) \overline{\text{year}}_{it})]^2}{\sum_{i=1}^N \sum_{t=1}^T [y_{it} - \bar{y}_{it}]^2}
\]

This gives us the proportion of total variance in an item attributable
to stable interpersonal differences.

To calculate the second component, which is the variance attributable to
systematic intrapersonal change, we calculate \(V(C)\) as follows:

\[
V(C) = 1 - \frac{\sum_{i=1}^N \sum_{t=1}^T [y_{it} - (\hat{\beta}_0 + \hat{\alpha}_{0i} + (\hat{\beta}_1 + \hat{\alpha}_{1i}) \text{year}_{it})]^2}{\sum_{i=1}^N \sum_{t=1}^T [y_{it} - \bar{y}_{it}]^2} - V(D)
\]

This is the incremental proportion of variance accounted for when we
allow the model predictions to change over time for each person.

The third component, which is the proportion of variance attributable to
measurement error or transient fluctuations, is what's left, such that
these three numbers sum to 1. As we discussed above, this component is
not of core interest here.

Together, \(V(D)\) and \(V(C)\) reflect the total systematic variance of
the outcome that the LCAM can attribute to either differences between
people or linear intrapersonal change over the course of the panel. Our
principal measure of interest is the proportion of systematic variance
accounted for by intrapersonal change, which we calculate as

@@

We can think of \(\omega\) as a type of Intraclass Correlation
Coefficient focused only on the systematic components of the model while
ignoring residual variance that might be attributable to measurement
error or non-persisting change (see also the variance decomposition of
period and cohort by Vaisey and Lizardo 2016). The higher this
proportion is, the more systematic variance in responses is attributable
to intrapersonal change during the panel. From now on, we use \(\omega\)
to stand for \emph{the proportion of systematic variance attributable to
intrapersonal change}.

Because this quotient summarizes variance proportions and therefore has
no natural referent, it is hard to say at which values we observe ``a
lot'' of intrapersonal change. But summarizing questions in this way
allows us to compare the relative importance of our two broad processes
across a range of questions that might have different levels of
non-systematic variance. This gives us a sense of the relative
prevalence of these two processes for each question, which is not
achievable under the tournament of models approach. This also allows us
to compare which kinds of questions show more or less intrapersonal
change relative to interpersonal differences. While the tournament of
models approach allowed researchers to say whether questions in general
showed \emph{any} evidence of updating, this approach allows us to make
more specific claims about the prevalence of these theoretical processes
for particular questions relative to each other.
